---
title: "Does Urban Design Predict </br>Sympathy for the Far Right?"
subtitle: "A Case Study of Vox and 73 Spanish Cities"
author: "Jonathan Kent | March 23, 2021"
output:
  xaringan::moon_reader:
    css: 
      - xaringan-themer.css
      - xaringan-title.css
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'
---

class: full-slide-fig

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  fig.asp = 9/16,
  fig.align = 'center',
  echo = F,
  message = F,
  warning = F,
  out.width = "85%",
  dpi= 300
)
```

```{r xaringan-themer}
library(xaringanthemer)
style_mono_accent(
  base_color = "#353889",
  header_font_google = google_font("Lora"),
  text_font_google   = google_font("Lato", 300),
  text_font_size = "1.4rem",
  title_slide_background_image = "demosocBG.png",
  text_color = "#05050d"
)
```

```{r xaringan-extra}
library(xaringanExtra)
use_logo(
  image_url = "https://waronimmigrants.github.io/images/demosoc_logo.png",
  width = "242px",
  height = "39px",
  position = css_position(top = "1em", right = "1em")
)
```

## Theoretical Framework

- Nativism is central to the modern far right, and it's ideology holds that race, gender, and other inequalities are natural and even positive (Mudde 2007, 2019).

- Contact theory research has demonstrated that such prejudicial views toward out-groups can be reduced though interaction (Pettigrew et al. 2011).

- Certain forms of urban design promote more contact between neighbors than others (Jacobs 1961, Gehl 2011).

- Implication: In some cities there should be more interaction, less prejudice, and less support for the far right.

---

## Framework: Far Right

- We are now in the fourth wave of the rise of the far right in Europe (Mudde 2019).

- At the individual level, research explaining this rise has focused on economic and cultural grievance hypotheses, with mixed results (Golder 2016).

- These same hypotheses have been applied to the social context, operationalized as unemployment rates and migrant populations (e.g.  Georgiadou et al. 2018).

- Most of the literature stays within this competition framework, though there are interesting exceptions (e.g. Bolet 2021).

---

## Framework: Contact Theory

- There is a very large literature that direct contact reduces prejudice (Pettigrew et al. 2011).

- However, newer literature on </i>extended</i> contact is particularly interesting (Zhou et al. 2018).

- People whose friends have out-group friends are less likely to be prejudiced.

- People who observe, but don't participate in, intergroup contact are also less likely to be prejudiced.

- At the scale of a city, the potential impact is enormous.

---

## Framework: Urban Theory

- Modernist design trends inserted openness into previously continuous urban fabric throughout the 20th century.

- Two dueling philosophies argued that cities should expand horizontally (urban sprawl) or vertically (set-back residential high-rises). For both, grass and light was essential.

- Jacobs (1961) famously sounded the alarm that these trends were degrading the vitality of cities, reducing interaction between neighbors.

- EEA's CORINE Land Cover data distinguishes between </i>continuous urban fabric</i> and </i>discontinuous urban fabric</i> which broadly captures the dichotomy between compact, traditional designs and modernist, open ones.

---

## Hypotheses

- H1: Residents of cities with more continuous urban fabric will be less likely to have sympathy for far-right parties.

- H2: The negative relationship between continuous urban fabric and far-right sympathy will be stronger in cities with larger relative migrant populations.

---

## Case Selection & Data

- I study a single country, to reduce variation in political context and maximize data availability and consistency.

- Vox is very similar to other radical-right parties in Europe, especially in its anti-immigrant and anti-feminism stances (Gould 2019, Arcila Calderón et al. 2020, Bernardez-Rodal et al. 2020).

- At the individual level, data comes from 12 pooled CIS barometer surveys from January 2019 to March 2020, which ask respondents how likely they are to support Vox on a 0 (never) to 10 (always) scale. The dichotomous outcome variable treats response 1-10 as *sympathy for the far right*.

- City-level data is provided by Eurostat.

---

## Methods

- Hypotheses are tested using Bayesian multilevel logistic regression models fitted with rstanarm.

- Models include urban fabric, population, and density, as well as variables common to the literature: unemployment rate, migrant population, and change in migrant population. All city-level variables are coded as one-standard-deviation z-scores.

- At the individual level, I include age, age<sup>2</sup>, gender, religion, education, and occupation variables. I model separate intercepts for all categorical variables.

- I also account for variation in the political context by including survey date as a continuous variable and CCAA as a level.

---

## Findings: Models

```{r}
library(tidyverse)
library(rstanarm)
library(bayesplot)
library(ggpubr)
library(plotly)

bayesplot_theme_set(theme_xaringan(background_color = "#FFFFFF"))
theme_set(theme_xaringan(background_color = "#FFFFFF",
                         title_font_size = 17,
                         text_font_size = 13,
                         text_color = "#05050d",
                         accent_color = "#0f1029",
                         text_font = "Lato",
                         text_font_use_google = TRUE,
                         title_font = "Lato",
                         title_font_use_google = TRUE))

load("~/Google Drive/PhD/Paper Two/M1.rda")
model <- M1
rm(M1)

load("~/Google Drive/PhD/Paper Two/M2.rda")
model_change <- M2
rm(M2)

load("~/Google Drive/PhD/Paper Two/RC1.rda")
model_density <- RC1
rm(RC1)

load("~/Google Drive/PhD/Paper Two/M3.rda")
model_interaction <- M3
rm(M3)

p1 <- mcmc_areas(as.matrix(model), pars = c("z.urban.fabric","z.migrants","z.unemployment","z.population"),prob = 0.9, prob_outer = 0.99) + geom_vline(xintercept = 0) + theme_xaringan(background_color = "#FFFFFF",
                         title_font_size = 13,
                         text_font_size = 9,
                         text_color = "#05050d",
                         accent_color = "#0f1029",
                         text_font = "Lato",
                         text_font_use_google = TRUE,
                         title_font = "Lato",
                         title_font_use_google = TRUE) + scale_y_discrete(labels=c("z.urban.fabric" = "Urban Fabric","z.population" = "Overall Pop.","z.migrants" =  "Migrant Pop.","z.unemployment" = "Unemployment"), limits = c("z.population","z.migrants","z.unemployment","z.urban.fabric"))

p2 <- mcmc_areas(as.matrix(model_change), pars = c("z.urban.fabric","z.migrants.change","z.unemployment","z.population"),prob = 0.9, prob_outer = 0.99)+ geom_vline(xintercept = 0) + theme_xaringan(background_color = "#FFFFFF",
                         title_font_size = 13,
                         text_font_size = 9,
                         text_color = "#05050d",
                         accent_color = "#0f1029",
                         text_font = "Lato",
                         text_font_use_google = TRUE,
                         title_font = "Lato",
                         title_font_use_google = TRUE) + scale_y_discrete(labels=c("z.urban.fabric" = "Urban Fabric","z.population" = "Overall Pop.","z.migrants.change" =  "Migrant Change","z.unemployment" = "Unemployment"), limits = c("z.population","z.migrants.change","z.unemployment","z.urban.fabric"))

p3 <- mcmc_areas(as.matrix(model_density), pars = c("z.urban.fabric","z.migrants","z.unemployment","z.population", "z.density"),prob = 0.9, prob_outer = 0.99) + geom_vline(xintercept = 0) + theme_xaringan(background_color = "#FFFFFF",
                         title_font_size = 13,
                         text_font_size = 9,
                         text_color = "#05050d",
                         accent_color = "#0f1029",
                         text_font = "Lato",
                         text_font_use_google = TRUE,
                         title_font = "Lato",
                         title_font_use_google = TRUE) + scale_y_discrete(labels=c("z.urban.fabric" = "Urban Fabric","z.population" = "Overall Pop.","z.migrants" =  "Migrant Pop.","z.unemployment" = "Unemployment", "z.density" = "Density"), limits = c("z.density","z.population","z.migrants","z.unemployment","z.urban.fabric"))

p4 <- mcmc_areas(as.matrix(model_interaction), pars = c("z.urban.fabric","z.migrants","z.unemployment","z.population", "z.urban.fabric:z.migrants"),prob = 0.9, prob_outer = 0.99) + geom_vline(xintercept = 0) + theme_xaringan(background_color = "#FFFFFF",
                         title_font_size = 13,
                         text_font_size = 9,
                         text_color = "#05050d",
                         accent_color = "#0f1029",
                         text_font = "Lato",
                         text_font_use_google = TRUE,
                         title_font = "Lato",
                         title_font_use_google = TRUE) + scale_y_discrete(labels=c("z.urban.fabric" = "Urban Fabric","z.population" = "Overall Pop.","z.migrants" =  "Migrant Pop.","z.unemployment" = "Unemployment", "z.urban.fabric:z.migrants" = "Urban Fab. x Mig. Pop."), limits = c("z.urban.fabric:z.migrants","z.population","z.migrants","z.unemployment","z.urban.fabric"))

ggarrange(p1, p2, p3, p4,
          labels = c("M1", "M2", "M3", "M4"),
          ncol = 2, nrow = 2)
```

---

## Findings: H1 Simulations

``` {r}
h1sim <- read_csv("~/Google Drive/PhD/Paper Two/h1sim.csv") %>% rename(Urban.Fabric = uf, Vox.Sympathy = vox)

ggplot(h1sim, aes(Urban.Fabric,Vox.Sympathy)) + geom_point(color = "#353889", alpha = .2, size = 3) + labs(x = "Counterfactual Continuous Urban Fabric (z-score)", y = "Sympathy for Vox (%)")
```

---

## Findings: H2 Simulations

``` {r echo = F, message = F}
h2sim <- read_csv("~/Google Drive/PhD/Paper Two/h2sim.csv") %>% rename(Urban.Fabric = uf, Vox.Sympathy = vox, Migrant.Pop = mig)

ggplot(h2sim, aes(Urban.Fabric, Vox.Sympathy, color = Migrant.Pop)) + geom_point(alpha = .15, size = 2) + geom_smooth(method = "lm") + theme(legend.position="bottom") + labs(x = "Counterfactual Continuous Urban Fabric (z-score)", y = "Sympathy for Vox (%)", color = "Migrant Pop.") + scale_color_manual(values = c("High (2)" = "#353889", "Low (-2)" = "#893538"))
```

---

## Sensitivity Analysis

``` {r echo = F, message = F}
cutoffs <- read_csv("~/research/stan_bi_runs.csv") %>% rename(pi10 = pi25, pi90 = pi75)
cutoffs$cutoff <- paste0("> ",cutoffs$cutoff)

ggplot() + geom_linerange(data = cutoffs, aes(y = cutoff, xmin = pi2_5, xmax = pi97_5), color = "#aeafcf", size = 2) + geom_linerange(data = cutoffs, aes(y = cutoff, xmin = pi10, xmax = pi90), color = "#353889", size = 2) + geom_vline(xintercept = 0) + labs(x = "Cont. Urban Fabric Coefficient", y = 'Coding of "Sympathy for Vox"')
```

---

## Conclusions

- The explanatory literature on far-right support should shift its focus from the economic grievance vs. cultural grievance debate. As this paper and Bolet (2021) show, there is much more to the social context.

- We may focus too much on immigration issues, as well. The far right seeks to perpetuate many other inequalities, especially gender.

- Social scientists should give more consideration to urban fabric. Climate change will force us to redesign our cities again, and we need to be part of the conversation.


---

## References
<font size="2">

Bernardez-Rodal, A., Rey, P.R., and Franco, Y.G. (2020). Radical right parties and anti-feminist speech on Instagram: Vox and the 2019 Spanish general election. <i>Party Politics</i>.</br>

Bolet, D. (2021). Drinking alone: Local socio-cultural degradation and radical right support, the case of British pub closures. <i>Comparative Political Studies</i>.</br>

Calderón, C. A., de la Vega, G., and Herrero, D. B. (2020). Topic modeling and char. of hate speech against immigrants on Twitter around the emergence of a far-right party in Spain. <i>Social Sciences</i>.</br>

Gehl, J. (2011). <i>Life between buildings: using public space</i>. Island Press, Washington, DC.</br>

Georgiadou, V., Rori, L., and Roumanias, C. (2018). Mapping the european far right in the 21st century:  A meso-level analysis. <i>Electoral Studies</i>, 54:103–115.</br>

Golder, M. (2016). Far right parties in Europe. <i>Annual Review of Political Science</i>, 19(1):477–497.</br>

Gould, R. (2019). Vox España and Alternative für Deutschland: Propagating the crisis of national identity. <i>Genealogy</i>, 3(4):64.</br>

Jacobs, J. (1961). <i>The  Death  and  Life  of  Great  American  Cities</i>. Random House, New York, NY.</br>

Mudde, C. (2007). <i>Populist radical right parties in Europe</i>. Cambridge University Press, Cambridge, UK.</br>

Mudde, C. (2019). <i>The far right today</i>. Polity, Cambridge, UK.</br>

Pettigrew, T.F., Tropp, L.R., Wagner, U., and Christ, O. (2011). Recent advances in intergroup contact theory. <i>International Journal of Intercultural Relations</i>, 35(3):271–280.</br>

Zhou, S., Page-Gould, E., Aron, A., Moyer, A., and Hewstone, M. (2018). The extended contact hypothesis: A meta-analysis on 20 years of research. <i>Personality and Social Psychology Review</i>.</br>

</font>